---
title: "Titanic Survival Analysis"
author: "MD.Mehedi,
        Tanjena Akter Shefa,
        Shahriyar Hoque Fihal"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ***Introduction:***

This report explores the survival patterns of passengers on the Titanic to determine whether key demographic and socio-economic factors influenced a person's likelihood of survival. Specifically, we construct contingency tables, generate bar charts, and perform Chi-square tests of independence to evaluate the impact of passenger sex and ticket class on survival outcomes.

## ***SECTION 1: SETUP & DATA PREPARATION***

## *Load the built-in Titanic dataset and convert it to a dataframe*

```{r}
titanic_df <- as.data.frame(Titanic)
titanic_data <- titanic_df[rep(1:nrow(titanic_df), titanic_df$Freq), -5]
rownames(titanic_data) <- NULL
```

## ***Section 2: Survival Analysis by sex***

### ***2.1:Contingency Table:***

The contingency table below shows the raw counts of survival broken down by gender.

```{r}
sex_table <- table(titanic_data$Sex, titanic_data$Survived)
sex_table

```

### ***2.2: Visualisation:***

The grouped bar plot below visually highlights the distribution of survivors and non-survivors across both sexes using base R graphics.

```{r}
plot_sex_table <- table(titanic_data$Survived, titanic_data$Sex)
barplot(plot_sex_table, 
        beside = TRUE, 
        col = c("red", 'blue'), 
        legend.text = c("No", "Yes"),
        args.legend = list(x = "topright", title = "Survived"),
        main = "Titanic Survival by Sex",
        xlab = "Sex", 
        ylab = "Passenger Count")
       
```

### ***2.3: Chi-square Test of independence:***

We test the null hypothesis ( Survival and Sex are completely independent) against the alternative hypothesis (Survival and Sex are dependent).

```{r}
chisq.test(sex_table)
```

Because the resulting p-value is extremely small . We reject the null hypothesis. There is strong, statistically significant evidence that survival on the Titanic was dependent on a passenger's sex.

## ***Section 3: Survival Analysis by passenger class:***

### ***3.1: Contingency Table:***

The table below displays survival counts across the different ticket classes (1st, 2nd, 3rd, and Crew).

```{r}
class_table <- table(titanic_data$Class, titanic_data$Survived)
class_table
```

### ***3.2: Visualization:***

The following chart compares the number of survivors and non-survivors across all passenger classes.

```{r}
plot_class_table <- table(titanic_data$Survived, titanic_data$Class)

barplot(plot_class_table, 
        beside = TRUE, 
        col = c("red", "blue"), 
        legend.text = c("No", "Yes"),
        args.legend = list(x = "topleft", title = "Survived"),
        main = "Titanic Survival by Passenger Class",
        xlab = "Class", 
        ylab = "Passenger Count")
```

### ***3.3: Chi-square Test of independence:***

We test the null hypothesis (Survival and Passenger Class are independent) against the alternative hypothesis (Survival and Passenger Class are dependent).

```{r}
chisq.test(class_table)
```

Because the p-value is so small (p \< 0.05), the test confirms that survival wasn't just random. The ticket class heavily influenced the chances of making it off the ship alive.
